Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-14 Manuscript/proof_outline.tex
review spacing of code in the paper
remove some blue

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see
  Definition~\ref{def:semi});
\item {\color{blue}a definition of additive contents over semi-rings (see Definition~\ref{def:content});}
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us with a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. In addition, {\color{blue}Isabelle works with simple type theory, while Lean is an implementation of dependent theory.}

@@

We now come to our contribution. 
%which we will PR to mathlib :
{\color{blue}The extension theorem is a statement about extending a
  set-function on a product space to a (finite) measure, where the
  product space can come with an arbitraty index set.}  The next
definition covers the important concept of a projective family of
measures. In short, we define measures on any finite subset of indices
in a consistent way. Finite sets (of some type \leanline{ι}) are
formalized with \leanline{Finset ι}. % Such a \leanline{Finset} is
defined as a \leanline{Multiset} (a
%quotient type of the \leanline{List}-type with respect to permutations
%of the list) together with a proof that it contains no duplicates\todo{remove this implementation detail?}.

@@

\begin{proof}
 Since $\alpha$ is separable, there is a subset $\{x_1, x_2,...\}
 \subseteq_c \alpha$ with closure $\alpha$. By separability, for each
 $n\in \mathbb N$, $\alpha = \bigcup_{k=1}^\infty
 B_{\varepsilon_n}(x_k)$. Since $\mu$ is continuous from above in
 $\emptyset$ (see Lemma~\ref{P:stetigmass})
  \[
  0 = \mu\Big(\Big(\bigcup_{k=1}^\infty B_{\varepsilon_n}(x_k)\Big)^c
  \Big) = \lim_{N\to\infty} \mu\Big( \Big(\bigcup_{k=1}^N
  B_{\varepsilon_n}(x_k)\Big)^c\Big).
  \]
  Let $\delta_n\downarrow 0$ be summable with $\sum_{n=1}^\infty
  \delta_n = 1$ (e.g.\ $\delta_n = 2^{-n}$). Then, there is some
  $N_n\in\mathbb N$ with $\mu\Big( E\setminus \bigcup_{k=1}^{N_n}
  B_{\varepsilon_n}(x_k)\Big) < \varepsilon \delta_n$. Now take
  \[
    A := \bigcap_{n=1}^\infty \bigcup_{k=1}^{N_n}
    B_{\varepsilon_n}(x_k),
  \]
  which by construction is totally bounded (for $r>0$ take $n$ such
  that $\varepsilon_n < r$, such that $B_r(x_1),...,B_r(x_{N_n})$
  cover $A$), and therefore has a compact closure as noted
  {\color{blue} in the lines directly preceding the lemma}. In
  addition, by $\sigma$-sub-additivity of $\mu$,
  \begin{align*}
     \mu((\overline A)^c) & \leq \mu(A^c) = \mu\Big(
     \bigcup_{n=1}^\infty \Big(\Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)\Big) \\ & \leq
     \sum_{n=1}^\infty \mu\Big( \Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)< \varepsilon,
  \end{align*}
 and we are done.
\end{proof}

>>>>

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see
  Definition~\ref{def:semi});
\item a definition of additive contents over semi-rings (see Definition~\ref{def:content});
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us with a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. In addition, {\color{blue}Isabelle works with simple type theory, while Lean is an implementation of dependent theory.}

@@

We now come to our contribution. 
The extension theorem is a statement about extending a
set-function on a product space to a (finite) measure, where the
product space can come with an arbitraty index set.  The next
definition covers the important concept of a projective family of
measures. In short, we define measures on any finite subset of indices
in a consistent way. Finite sets (of some type \leanline{ι}) are
formalized with \leanline{Finset ι}. % Such a \leanline{Finset} is
defined as a \leanline{Multiset} (a
%quotient type of the \leanline{List}-type with respect to permutations
%of the list) together with a proof that it contains no duplicates\todo{remove this implementation detail?}.

@@

\begin{proof}
 Since $\alpha$ is separable, there is a subset $\{x_1, x_2,...\}
 \subseteq_c \alpha$ with closure $\alpha$. By separability, for each
 $n\in \mathbb N$, $\alpha = \bigcup_{k=1}^\infty
 B_{\varepsilon_n}(x_k)$. Since $\mu$ is continuous from above in
 $\emptyset$ (see Lemma~\ref{P:stetigmass})
  \[
  0 = \mu\Big(\Big(\bigcup_{k=1}^\infty B_{\varepsilon_n}(x_k)\Big)^c
  \Big) = \lim_{N\to\infty} \mu\Big( \Big(\bigcup_{k=1}^N
  B_{\varepsilon_n}(x_k)\Big)^c\Big).
  \]
  Let $\delta_n\downarrow 0$ be summable with $\sum_{n=1}^\infty
  \delta_n = 1$ (e.g.\ $\delta_n = 2^{-n}$). Then, there is some
  $N_n\in\mathbb N$ with $\mu\Big( E\setminus \bigcup_{k=1}^{N_n}
  B_{\varepsilon_n}(x_k)\Big) < \varepsilon \delta_n$. Now take
  \[
    A := \bigcap_{n=1}^\infty \bigcup_{k=1}^{N_n}
    B_{\varepsilon_n}(x_k),
  \]
  which by construction is totally bounded (for $r>0$ take $n$ such
  that $\varepsilon_n < r$, such that $B_r(x_1),...,B_r(x_{N_n})$
  cover $A$), and therefore has a compact closure as noted
  in the lines directly preceding the lemma. In
  addition, by $\sigma$-sub-additivity of $\mu$,
  \begin{align*}
     \mu((\overline A)^c) & \leq \mu(A^c) = \mu\Big(
     \bigcup_{n=1}^\infty \Big(\Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)\Big) \\ & \leq
     \sum_{n=1}^\infty \mu\Big( \Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)< \varepsilon,
  \end{align*}
 and we are done.
\end{proof}
